---
title: "[Zenith](https://github.com/GabrielHoffman/zenith) gene set testing after dream analysis"
subtitle: "Zenith controls the false positive rate while retaining power"
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)"
  affiliation: | 
    Icahn School of Medicine at Mount Sinai, New York
date: "`r date()`"
abstract: >
 Perform gene set analysis on the result of differential expression using linear (mixed) modeling with [dream](https://doi.org/btaa687) by considering the correlation between gene expression traits.  [Zenith](https://github.com/GabrielHoffman/zenith) performs competitive gene sets analysis by comparing the tests statistics from genes in a  gene set to the rest of the genes not in the set while modeling correlation between genes in the set.  This package is a slight modification of [limma::camera](https://doi.org/10.1093/nar/gks461) to 1) be compatible with dream, and 2) allow identification of gene sets with log fold changes with mixed sign by using the squared test statistic.  
vignette: >
  %\VignetteIndexEntry{Example usage of zenith on GEUVAIDIS RNA-seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
output:
  html_document:
    toc: true
    toc_float: true
---


```{r knitr.setup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)


# rmarkdown::render("geuvadis.Rmd")


# devtools::reload("/Users/gabrielhoffman/workspace/repos/zenith")

# scp geuvadis.html sklar1:/hpc/users/hoffmg01/www/software/zenith

```


## Load and process RNA-seq data
```{r load, message=FALSE, cache=FALSE}
library(recount)
library(variancePartition)
library(edgeR)
library(limma)
library(zenith)
library(ggplot2)
library(EnrichmentBrowser)
library(GSEABase)
library(dplyr)
library(kableExtra)

# For ENSEMBL id ENSG00000279457.4, return ENSG00000279457
trim_ensembl_ids = function(x){
  gsub("(.*)\\.(.*)", "\\1", x) 
}
```

#### Load RNA-seq data and metadata
```{r download}
# Download GEUVADIS data from recount2
if( ! file.exists("ERP001942/rse_gene.Rdata") ){
  download_study('ERP001942')
}

# Load the data
load(file.path('ERP001942', 'rse_gene.Rdata'))

# scale counts
rse <- scale_counts(rse_gene)

# get count data
geneCounts = assay(rse)

# read and format metadata
info = read.table("https://www.ebi.ac.uk/arrayexpress/files/E-GEUV-1/E-GEUV-1.sdrf.txt", header=TRUE, sep="\t")
colnames(info) = gsub("Comment\\.", "", colnames(info))
colnames(info) = gsub("Characteristics\\.", "", colnames(info))
colnames(info) = gsub("\\.$", "", colnames(info))

# extract unique samples and drop a problematic sample
info = unique(info[,!colnames(info) %in% c('Scan.Name', 'SUBMITTED_FILE_NAME', 'FASTQ_URI')])
info = info[info$ENA_RUN != 'ERR188328',]

# merge sample information in colData(rse) and info
metadata = merge(colData(rse), info, by.x="run", by.y='ENA_RUN')
metadata = data.frame(metadata)
rownames(metadata) = metadata$run

# keep RNA-seq samples in metadata
geneCounts = geneCounts[,colnames(geneCounts) %in% metadata$run]
geneCounts = geneCounts[,match(metadata$run, colnames(geneCounts))]
```

#### RNA-seq processing
```{r processing}
# keep only expressed genes
isexpr = rowSums(cpm(geneCounts)>1) >= 0.2*ncol(geneCounts)

# apply TMM normalization
dge = DGEList( geneCounts[isexpr,] )
dge = calcNormFactors(dge)

# Create log2 counts per million for plotting
geneExpr = cpm(dge, log=TRUE)

# extract gene annotations
geneInfo = data.frame(rowData(rse))
```

#### Confirm sex based on expression
Plot expression of XIST and UTY  
We see a clear separation of sex based on XIST from chrX and UTY from chrY.
```{r plot.sex, fig.height=5, fig.width=5}
# extract ENSEMBL identifies for XIST and UTY
ens_XIST = geneInfo$gene_id[which(geneInfo$symbol == "XIST")]
ens_UTY = geneInfo$gene_id[which(geneInfo$symbol == "UTY")]

df = data.frame(metadata, XIST= geneExpr[ens_XIST,],
        UTY = geneExpr[ens_UTY,])

ggplot(df, aes(XIST, UTY, color=sex)) + geom_point() + theme_bw() + theme(aspect.ratio=1) + scale_color_manual(values = c("red", "blue"))
```



#### Differential expression with `dream`
```{r diffExpr, fig.height=5, fig.width=5}
form = ~ sex + ancestry.category + laboratory
vobj = voomWithDreamWeights(dge, form, metadata, plot=TRUE)

fit = dream(vobj, form, metadata, computeResiduals=TRUE)
fit = eBayes(fit)
```

#### Load genesets
```{r load.genesets}
# Load Gene Ontology database 
go.gs = get_GeneOntology()
   
# convert from GeneSetCollection to list used by camera and zenith
geneSets_GO = recodeToList( go.gs )
```


## Zenith gene set analysis
```{r zenith}
# Map from Ensembl genes in geneSets_GO to 
# from trimmed Ensembl names from RNA-seq data 
index = ids2indices( geneSets_GO, trim_ensembl_ids(rownames(fit)))
   
# filter by size of gene set
n_genes_in = 10
index = index[sapply(index, length) >= n_genes_in]

# Run camera
dsgn = model.matrix( form, metadata)
res_camera = camera( vobj, index, dsgn, inter.gene.cor=NA, contrast="sexmale")

# Run zenith.  
# Note that zenith allows analysis of any dream model and 
# any coefficient or contrast defined in the model. 
res_zenith = zenith( fit, "sexmale", index)
```

### Compare camera and zenith

#### Results from camera
```{r ktable.camera}
res_camera[1:8,] %>% kable %>% kable_styling(full_width = FALSE) 
```

#### Results from zenith
```{r kable.zenith}
res_zenith[1:8,] %>% kable %>% kable_styling(full_width = FALSE)     
```

While there is a very slight difference in the results from the methods, this is attributable to the very slight difference in the correlation value computed.  This appears to be due to an numerical issue with `camera`, but values are correct to two digits.   
```{r compare, fig.height=5, fig.width=5}
df = merge(res_camera, res_zenith, by="row.names")
plot(df$Correlation.x, df$Correlation.y, xlab="Correlation from zenith", ylab="Correlation from camera")
abline(0, 1, col="red")
```

## Zenith with squared test statistic
Using the squared test statistic allows zenith to identify gene sets with both positive and negative effects.  But note that the `Direction` must be interpreted differently here.  `Up` indicates that the square test statistics from genes in the gene set are larger then the background.  Because the test statistics are squared, the effects can be a mix or positive and negative.  Therefore `Up` is the relevant test here.  Conversely, `Down` indicates that the genes in the gene set are __less__ associated with the variable than the background.  So gene sets that are `Down` are not usually of biological interest.
```{r zenith_squared}
res_zenith2 = zenith( fit, "sexmale", index, squaredStats=TRUE)    
```

#### Results from zenith
```{r kable.zenith_squared}
res_zenith2[1:20,] %>% kable %>% kable_styling(full_width = FALSE)    
```

## Analysis with MSigDB
#### Load [MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/index.jsp) and perform zenith analysis
```{r MSigDB}
gs_msigdb = get_MSigDB()

# Map from Ensembl genes in gs_msigdb to 
# from trimmed Ensembl names from RNA-seq data 
index = ids2indices( recodeToList(gs_msigdb), trim_ensembl_ids(rownames(fit)))

# filter by size of gene set
n_genes_in = 10
index = index[sapply(index, length) >= n_genes_in]

 # zenith analysis of MSigDB using square test statistics
res_zenith_msigdb = zenith( fit, "sexmale", index, squaredStats=TRUE)
```

```{r kable.zenith_squared.msigdb}
res_zenith_msigdb[1:20,] %>% kable %>% kable_styling(full_width = FALSE)
```

#### Zenith controls the false positive rate 
```{r testnull}

set.seed(1)
# simulate random variable
metadata$rnd = rnorm(nrow(metadata))

# fit dream model
form = ~ sex + ancestry.category + laboratory + rnd

fit = dream(vobj, form, metadata, computeResiduals=TRUE)
fit = eBayes(fit)

# fit zenith model, testing the coefficent "rnd"
res_zenith = zenith( fit, "rnd", index)

# fit zenith model for squared statistics, testing the coefficent "rnd"
res_zenith_sq = zenith( fit, "rnd", index, squaredStats=TRUE)
```

index = lapply( 1:2000, function(x){
  sample.int(nrow(fit), 100)
  })
names(index) = paste0("gs_", 1:length(index))

res1 = zenith( fit, "rnd", index, use.ranks=FALSE, allow.neg.cor=TRUE)

# fit zenith model for squared statistics, testing the coefficent "rnd"
res2 = zenith( fit, "rnd", index, squaredStats=TRUE, use.ranks=FALSE, allow.neg.cor=TRUE)

res3 = cameraPR( topTable(fit, "rnd", sort.by="none", number=Inf)$t, index, inter.gene.cor=0.2)

sum(res1$PValue < 0.05) / nrow(res1)
sum(res2$PValue < 0.05) / nrow(res2)
sum(res3$PValue < 0.05) / nrow(res3)


par(mfrow=c(1,4))
hist(res1$Correlation)
hist(res1$PValue)
hist(res1$p.less)
hist(res1$p.greater)


hist(res2$Correlation)
hist(res2$PValue)
hist(res2$p.less)
hist(res2$p.greater)


# hist(res2$delta) 
hist(res3$PValue, main="CameraPR")
hist( 2*pmin(res2[names(index),]$p.less, res2[names(index),]$p.greater), main="Two sided zenith" )


plot(res3[names(index),]$PValue, 2*pmin(res2[names(index),]$p.less, res2[names(index),]$p.greater))


cameraPR( topTable(fit, "rnd", sort.by="none", number=Inf)$t^2, index[760], inter.gene.cor=0.002167243)

Stat = topTable(fit, "rnd", sort.by="none", number=Inf)$t

res = lapply( index, function(iset){
  zenith:::.rankSumTestWithCorrelation(iset, statistics=Stat^2, correlation=0.8, df=Inf)
})
res = do.call(rbind, res)


hist(res$less)
hist(res$greater)



I have show that 
1) zenith controls the false positive rate under the null when using random variable in the dream model
2) You can see in the vignette that the squareStats give much more findings that the up/down analysis: https://hoffmg01.u.hpc.mssm.edu/software/zenith/geuvadis.html

3) The p-values are uniform, when variable is random and there is no correlation structure in the data


In my testing experience, t^2 is succesptible to outliers driving the association.
  I can use abs(t) instead while still using rho^2 to approximate the correlation

# devtools::reload("/Users/gabrielhoffman/workspace/repos/zenith")



with use.rank = TRUE, statistic is always conservative
  But centered around 0.5 as corr -> 1

with use.rank = FALSE, delta is affected by outliers and is not symmetric
  can I used abs(t) instead? since it is less prone to outliers

  if tstats are correlated rho, how are abs(tstats) correlated?


library(mvtnorm)

df = lapply(seq(0, 1, length.out=1000), function(rho){
  C = matrix(rho, 2,2)
  diag(C) = 1
  X = rmvnorm(10000, sigma=C)

  data.frame(rho, 
    rho_hat = cor(X[,1], X[,2]), 
    rho_sq = cor(X[,1]^2, X[,2]^2), 
    rho_abs = cor(abs(X[,1]), abs(X[,2])))
})
df = do.call(rbind, df)

par(mfrow=c(1,4))
plot(df$rho, df$rho_hat)
plot(df$rho, df$rho_sq)
plot(df$rho, df$rho_abs)
plot(df$rho_sq, df$rho_abs)
abline(0, 1, col="red", lwd='2')

rho = seq(0, 1, length.out=1000)
plot(rho, rho^2 )











